Normalization by Evaluation for Sized Dependent Types

### 2017-02-27 nbe/icfp17.tex
[ ICFP 17 ] Moved Sec. on type checking to long version.

In this section, we show that type checking for normal forms is decidable, and succeeds for those which can be type via the restricted rule for size polymorphism elimination:
\[
  \ru{\Gamma \ders t : \forall T \qquad
      \resurrect\Gamma \der a : \Size
    }{\Gamma \ders t\, \ann a : T[a]}
\]
We will refer to the restricted typing judgement as \fbox{$\Gamma \ders t : T$}, and obviously,
if\/ $\Gamma \ders t : T$ then $\Gamma \der t : T$.

Bidirectional typing of normal forms.

Checking \fbox{$\Gamma \der t \jchk T$}.
\begin{gather*}
  \ru{T' \evalsto s \qquad
      \Gamma \der a : \Size
    }{\Gamma \der \Nat a \jchk T'}
\qquad
  \ru{T' \evalsto \Set[\ell'] \qquad \ell < \ell'
    }{\Gamma \der \Set[\ell] \jchk T'}
\\[2ex]
  \ru{T' \evalsto s \qquad
      \Gamma \der U \jchk s \qquad
      \cext \Gamma U \der T \jchk s
    }{\Gamma \der \piT U T \jchk T'}
\qquad
  \ru{T' \evalsto s \qquad
      \cext \Gamma \Size \der T \jchk s
    }{\Gamma \der \epiT \Size T \jchk T'}
%\\[2ex]
%   \ru{T' \evalsto \piT U T \qquad
%       \cext \Gamma U \der t \jchk T
%     }{\Gamma \der \lambda t \jchk T'}
% \qquad
%   \ru{T' \evalsto \epiT \Size T \qquad
%       \eext \Gamma \Size \der t \jchk T
%     }{\Gamma \der \lambda t \jchk T'}
\\[2ex]
  \ru{T' \evalsto \Nat b \qquad
      \resurrect \Gamma \der a+1 \leq b : \Size
    }{\Gamma \der \zero a \jchk T'}
\qquad
  \ru{T' \evalsto \Nat b \qquad
      \resurrect \Gamma \der a+1 \leq b : \Size \qquad
      \Gamma \der t \jchk \Nat a
    }{\Gamma \der \suc a t \jchk T'}
\\[2ex]
  \ru{T' \evalsto \epiT U T \qquad
      \eext \Gamma U \der t \jchk T
    }{\Gamma \der \lambda t \jchk T'}
\qquad
  \ru{\Gamma \der t \jinf T \qquad \Gamma \der T \asub T'
    }{\Gamma \der t \jchk T'}
\end{gather*}

Inference \fbox{$\Gamma \der t \jinf T$}.
\begin{gather*}
  \ru{\Gamma(i) = \erann[\noterased] T
    }{\Gamma \der \ind i \jinf T}
\qquad
  \ru{\Gamma \der t \jinf T' \qquad
      T' \evalsto \piT U T \qquad
      \Gamma \der u \jchk U
    }{\Gamma \der t\, u \jinf T[u]}
\\[2ex]
  \ru{\Gamma \der t \jinf T' \qquad
      T' \evalsto \piT \Size T \qquad
      \Gamma \der a : \Size
    }{\Gamma \der t\,a \jinf T[a]}
\qquad
  \ru{\Gamma \der t \jinf T' \qquad
      T' \evalsto \erpiT \Size T \qquad
      \resurrect\Gamma \der a : \Size
    }{\Gamma \der t\,\ann a \jinf T[a]}
\\[2ex]
  \ru{\Gamma \der u \jinf \Nat (a+1) \qquad
      \Gamma \der T \jchk \Nat (a+1) \to \Set[\ell] \qquad
      \Gamma \der t_z \jchk T\,(\zero a) \qquad
      \Gamma \der t_s \jchk \funT x {\Nat a} {T\,(\suc a x)}
    }{\Gamma \der u\, \case[\ell] T {t_z} {t_s} \jinf T\,u}
\\[2ex]
  \ru{\Gamma \der u \jinf \Nat a \qquad
      \Gamma \der T \jchk \FixK\,\ell \qquad
      \Gamma \der t \jchk \FixT\,T
    }{\Gamma \der u\, \fix[\ell] T t \jinf T\,a\,u}
\end{gather*}

We could go beyond normal forms, by adding inference rules for the $\Nat$-constructors:
\begin{gather*}
  \ru{\resurrect \Gamma \der a : \Size
    }{\Gamma \der \zero a \jinf \Nat(a+1)}
\qquad
  \ru{\resurrect \Gamma \der a : \Size \qquad
      \Gamma \der t \jchk \Nat a
    }{\Gamma \der \suc a t \jinf \Nat(a+1)}
\end{gather*}

\begin{theorem}[Soundness of type checking]
  Let $\der \Gamma$.
  \begin{enumerate}
  \item If\/ $\resurrect\Gamma \der T$ and $\DD :: \Gamma \der t \jchk T$ then $\Gamma \ders t : T$.
  \item If\/ $\DD :: \Gamma \der t \jinf T$ then $\resurrect\Gamma \der T$ and $\Gamma \ders t : T$.
  \end{enumerate}
\end{theorem}
\begin{proof}
  Simultaneously by induction on $\DD$, using subject reduction (Thm.~\ref{thm:sr})
  and soundness of algorithmic subtyping (Lemma~\ref{lem:asubsound}).
\end{proof}

\begin{lemma}[Weak head reduction of subtypes]
\label{lem:whdsub}
Let $\DD :: \Gamma \der T \asub T'$.
\begin{enumerate}
\item If\/ $T' \evalsto \Nat a'$ then $T \evalsto \Nat a$ and $\Gamma \der a \asub a' : \Size$.
\item If\/ $T' \evalsto \Set[\ell']$ then $T \evalsto \Set[\ell]$ and $\ell \asub \ell'$.
\item If\/ $T' \evalsto \piT{A'}{B'}$ then $T \evalsto \piT {A} {B}$ and
      $\Gamma \der A' \asub A$ and $\cext \Gamma {A'} \der B \asub B'$.
\item If\/ $T' \evalsto \epiT \Size {B'}$ and $T \evalsto \epiT \Size B$ and $\cext \Gamma \Size \der B \asub B'$.
\end{enumerate}
\end{lemma}
\begin{proof}
  By cases on $\DD$, since weak head evaluation is deterministic.
\end{proof}
\begin{lemma}[Weak head reduction of supertypes]
\label{lem:whdsuper}
Let $\DD :: \Gamma \der T \asub T'$.
\begin{enumerate}
\item If\/ $T \evalsto \Nat a$ then $T' \evalsto \Nat a'$ and $\Gamma \der a \asub a' : \Size$.
\item If\/ $T \evalsto \Set[\ell]$ then $T' \evalsto \Set[\ell']$ and $\ell \asub \ell'$.
\item If\/ $T \evalsto \piT A B$ then $T' \evalsto \piT {A'} {B'}$ and
      $\Gamma \der A' \asub A$ and $\cext \Gamma {A'} \der B \asub B'$.
\item If\/ $T \evalsto \epiT \Size B$ and $T' \evalsto \epiT \Size B'$ and $\cext \Gamma \Size \der B \asub B'$.
\end{enumerate}
\end{lemma}
\begin{proof}
  By cases on $\DD$, since weak head evaluation is deterministic.
\end{proof}
\begin{lemma}[Subsumption for type checking]
  \label{lem:subtc}
  Let $\sid : \Gamma' \leq \Gamma$.
  \begin{enumerate}
  \item If\/ $\DD :: \Gamma \der t \jchk T$ and $\resurrect\Gamma \der T \leq T'$ then $\Gamma' \der t \jchk T'$.
  \item If\/ $\DD :: \Gamma \der t \jinf T$ then $\Gamma' \der t \jinf T'$ and $\resurrect{\Gamma'} \der T \leq T'$.
  \end{enumerate}
\end{lemma}
\begin{proof}
  Simultaneously by induction on $\DD$, using % weak head normalization of types,
  lemmata \ref{lem:whdsub} and \ref{lem:whdsuper},
  and soundness and completeness of algorithmic subtyping.
\end{proof}
\begin{theorem}[Completeness of type checking for normal terms]
\bla
\begin{enumerate}
\item If\/ $\DD :: \Gamma \ders v : T$ then $\Gamma \der v \jchk T$.
\item If\/ $\DD :: \Gamma \ders m : T$ then $\Gamma \der m \jinf U$ and $\resurrect\Gamma \der U \leq T$.
\end{enumerate}
\end{theorem}
\begin{proof}
Simultaneously by induction on $\DD$, using (strong) inversion and Lemma~\ref{lem:subtc}.
\end{proof}

\begin{lemma}[Termination of type checking]
  Let $\der \Gamma$.
  \begin{enumerate}
  \item The query $\Gamma \der t \jinf \mbox{?}$ terminates.
  \item If\/ $\resurrect\Gamma \der T$ then the query $\Gamma \der t \jchk T$ terminates.
  \end{enumerate}
\end{lemma}
\begin{proof}
  By induction on $t$, using type weak head normalization and soundness of type checking, to maintain well-formedness of types.
  And, of course, decidability of subtyping.
\end{proof}

\begin{theorem}[Decidability of type checking for normal terms]
  Let $\der \Gamma$ and $\resurrect \Gamma \der T$.
  Then $\Gamma \ders v : T$ is decided by $\Gamma \der v \jchk T$.
\end{theorem}

>>>>

\LONGVERSION{
\input{sec-tycheck}
} % END LONG VERSION
\SHORTVERSION{
It is now straightforward to define a bidirectional type-checking
algorithm which decides typing for normal forms, and succeeds for
those that rely only on the following restricted rule for irrelevant
size application:
\[
  \ru{\Gamma \ders t : \forall T \qquad
      \resurrect\Gamma \der a : \Size
    }{\Gamma \ders t\, \ann a : T[a]}
\]
For lack of space, we delegate definition and soundness proof to the
long version of this article.
}
